---
title: "HPDM208Z Stratified Medicine Week 4 Workshop: Medications Part 2"
subtitle: "MSc Health Data Science, University of Exeter"
author: 
 - name: "Dr Luke C Pilling"
   email: "L.Pilling@exeter.ac.uk"
date: "2026-01-15"
date-format: iso
title-block-banner: true
format: html
number-sections: true
toc: true
knitr:
  opts_chunk: 
    R.options:
      width: 120

# Extension configuration https://github.com/mcanouil/quarto-collapse-output
filters:
  - collapse-output
---

```{r, include=FALSE}
options(tibble.width = 200)
ggplot2::theme_set(ggplot2::theme_bw(base_size = 12))
```


# Introduction

In this workshop you will extract prescriptions that contain two common antihypertensive medications: amlodipine (a calcium channel blocker [CCB]) and ramipril (an ACE inhibitor [ACEi]). You will perform epidemiological analysis with the derived data, identifying patient characteristics associated with being prescribed one or the other. 

This builds on skills from previous modules, and the workshops you have already undertaken. Refer to previous workshops for Linux and R code to complete this workshop.

We'll use the same synthetic version of UK Biobank used is Part 1. The data is synthetic to protect participant anonymity, though it is designed to resemble real UK Biobank data.


# Setup

## OpenStack

Log in to your OpenStack instance as usual. Activate the `conda` environment by typing the below into your Linux terminal.

```{bash, eval=FALSE}
conda activate hpdm208z
```

We will use some linux command line tools you have met before for file manipulation, and then R for statistical analysis.

We'll use functions from the {[tidyverse](https://tidyverse.org/)} suite of packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures, providing a comprehensive set of tools for data manipulation, analysis, and visualisation.

If not already installed in your `conda` environment, type the below into your Linux terminal.

```{bash, eval=FALSE}
conda install -c r r-tidyverse -y
```

You can complete this workshop directly in the linux terminal, or create a JupyterLab notebook if you prefer. If the former make sure to save your commands in a text file.

## Working directory

All data for this workshop are on your OpenStack instance at the below location. Code examples assume this is your working directory:

```{bash, eval=FALSE}
~/hpdm208z/workshops/medications_data/
```


# UK Biobank GP prescribing data

This synthetic version of primary care (GP) prescribing data linked to UK Biobank can be viewed as below:

```{bash}
zcat ukb_gp_scripts.tsv.gz | head -n15
```

It is a tab-separated file, where each row is a prescribing event. Each person can have multiple rows, like with the clinical events data. In this synthetic dataset all prescriptions are coded with British National Formulary (`bnf`) codes. Real data may include other code types, such as SNOMED CT.


# Identify BNF clinical codes for amlodipine and ramipril

Identify the BNF Chemical Substance code for each medication. 

When considering atorvastatin previously we knew that the products only contained the single chemical substance. Products for conditions such as hypertension can include multiple chemical substances in the same tablet. 

Search for amlodipine in the BNF file and inspect the matched rows. Amlodipine (a calcium channel blocker [CCB]) can be prescribed alongside Olmesartan medoxomil (an angiotension receptor blocker [ARB]), so we need to identify the Chemical Substance code that refers to procuts containing amlodipine only.

```{bash}
# get rows with amlodipine (note the -i 'ignore case' flag)
# keep all columns to inspect file output
zcat bnf_code_historic_2024_version_86.tsv.gz | grep -i "amlodipine" > bnf_amlodipine.tsv
cut -f10-13 bnf_amlodipine.tsv
```

Looks like it is just one code that is specific for products containing only amlodipine. Save that to a file to use when extracting the GP prescriptions. (Note you can also see brand names for amlodipine here such as "Istin" but it is the same chemical substance). 

```{bash}
echo "0206020A0" > bnf_amlodipine.tsv
```

Now repeat and find the BNF rows that include ramipril. Inspect the file and find the Chemical Substance code that refers to products containing only ramipril.

```{bash}
# get rows with ramipril (note the -i 'ignore case' flag)
# keep all columns to inspect file output
zcat bnf_code_historic_2024_version_86.tsv.gz | grep -i "ramipril" > bnf_ramipril.tsv
cut -f10-13 bnf_ramipril.tsv
```

Looks like it is just one code that is specific for ramipril as well. Save that to a file to use when extracting the GP prescriptions.

```{bash}
echo "0205051R0" > bnf_ramipril.tsv
```


# Subset prescription records

Extract the rows in `ukb_gp_scripts.tsv.gz` that correspond to prescriptions containing the amlodipine or ramipril chemical substance only. Also include the file header, so that columns are labelled.

Use Linux command line tools (such as `grep`) to achieve this. 

```{bash}
# first get the header row
zcat ukb_gp_scripts.tsv.gz | head -n 1 > ukb_gp_scripts.amlodipine.tsv
zcat ukb_gp_scripts.tsv.gz | head -n 1 > ukb_gp_scripts.ramipril.tsv

# next search the GP scripts file for codes contained in a reference file
# we can do this in several ways. We can use grep directly, or use tools such as `awk`
# for BNF we do not want exact matches: we want to match any rows with the prefix
zgrep --fixed-strings --file bnf_amlodipine.tsv ukb_gp_scripts.tsv.gz >> ukb_gp_scripts.amlodipine.tsv
zgrep --fixed-strings --file bnf_ramipril.tsv ukb_gp_scripts.tsv.gz >> ukb_gp_scripts.ramipril.tsv
```

We are now ready to use R to load the (synthetic) UK Biobank baseline data, extracted GP prescriptions, and undertake some epidemiological analysis.


# Launch R and tidyverse

To launch R on the Linux command line just type `R`. The following commands in this workshop are designed to be run in R itself.

Load the {tidyverse} library.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Load UK Biobank data

## Baseline assessment data

Load and inspect the data. Subset to those with GP records available.

```{r, message=FALSE, warning=FALSE}
ukb <- read_tsv("ukb_baseline.tsv.gz")
```

```{r}
head(ukb)
ukb <- ukb |> filter(in_gp==1)
dim(ukb)
```

## GP prescriptions of antihypertensives

Load and inspect the extracted GP prescriptions data. Check it looks "OK".

We have kept the amlodipine and ramipril prescriptions separately as the moment, so that it is easy to create the "binary" variables later (such as `amlodipine_prevalent`).

Check your extracts have similar number of rows and structure to these.

### Amlodipine

```{r, message=FALSE, warning=FALSE}
ukb_scripts_amlodipine <- read_tsv("ukb_gp_scripts.amlodipine.tsv")
```

```{r}
dim(ukb_scripts_amlodipine)
head(ukb_scripts_amlodipine)
```

How many unique participants are there?

```{r}
length(unique(ukb_scripts_amlodipine$eid))
```

Does the data look sensible?

```{r}
summary(ukb_scripts_amlodipine$date)
```

### Ramipril

```{r, message=FALSE, warning=FALSE}
ukb_scripts_ramipril <- read_tsv("ukb_gp_scripts.ramipril.tsv")
```

```{r}
dim(ukb_scripts_ramipril)
head(ukb_scripts_ramipril)
```

How many unique participants are there?

```{r}
length(unique(ukb_scripts_ramipril$eid))
```

Does the data look sensible?

```{r}
summary(ukb_scripts_ramipril$date)
```


# Investigate participants prescribed antihypertensives

For some types of analysis such as longitudinal analysis, it is useful to keep the data in this "long" format (where participants have multiple rows). 

Today we just need to find the participants who have prescriptions before or after assessment, and investigate the participant characteristics associated with drug choice (CCB vs. ACEi).

## Identify prevalent prescriptions

Next let's work out which participants were prescribed amlodipine or ramipril before the baseline assessment (i.e., prevalent prescriptions), and which were prescribed after (incident prescriptions).

Join the assessment date with the prescription data.

```{r, message=FALSE, warning=FALSE}
ukb_scripts_amlodipine <- left_join(ukb_scripts_amlodipine, ukb[,c("eid","assessment_date")], by="eid")
ukb_scripts_ramipril   <- left_join(ukb_scripts_ramipril, ukb[,c("eid","assessment_date")], by="eid")
```

Create binary (0/1) variables indicating whether each prescription was before or after the UK Biobank assessment. **Hint**: use "if else" to assign [1] if a condition is met, otherwise assign [0] to the participants.

```{r}
ukb_scripts_amlodipine <- ukb_scripts_amlodipine |> mutate(
  amlodipine_prevalent = if_else(date <= assessment_date, 1, 0),
  amlodipine_incident  = if_else(date >  assessment_date, 1, 0)
)
table(ukb_scripts_amlodipine$amlodipine_prevalent)
```

```{r}
ukb_scripts_ramipril <- ukb_scripts_ramipril |> mutate(
  ramipril_prevalent = if_else(date <= assessment_date, 1, 0),
  ramipril_incident  = if_else(date >  assessment_date, 1, 0)
)
table(ukb_scripts_ramipril$ramipril_prevalent)
```

Now create corresponding binary variables in the main `ukb` data object, indicating whether a participant has a prevalent or incident prescription for each medication. **Hint**: use "if else" to assign [1] if a condition is met, otherwise assign [0] to the participants.

```{r}
ukb <- ukb |> dplyr::mutate(
  amlodipine_prevalent = if_else(eid %in% ukb_scripts_amlodipine$eid[ukb_scripts_amlodipine$amlodipine_prevalent==1], 1, 0),
  amlodipine_incident  = if_else(eid %in% ukb_scripts_amlodipine$eid[ukb_scripts_amlodipine$amlodipine_incident==1], 1, 0),
  ramipril_prevalent   = if_else(eid %in% ukb_scripts_ramipril$eid[ukb_scripts_ramipril$ramipril_prevalent==1], 1, 0),
  ramipril_incident    = if_else(eid %in% ukb_scripts_ramipril$eid[ukb_scripts_ramipril$ramipril_incident==1], 1, 0)
)
```

You should have numbers similar to these (note that some people have both prevalent and incident prescriptions!):

```{r}
table(ukb$amlodipine_prevalent)
table(ukb$amlodipine_incident)
table(ukb$ramipril_prevalent)
table(ukb$ramipril_incident)
```


## Create a binary variable `amlodipine_or_ramipril` 

This new variable will be useful for comparing predictors of drug choice. 

Create a new categorical variable called `amlodipine_or_ramipril`.  Set participants who were prescribed ramipril prior to baseline to `0` and amlodipine to `1`. If a participants received both prior to baseline set them to `2`. Other participants should be missing (i.e., set to `NA`). **Hint**: use "if else".

```{r}
ukb$amlodipine_or_ramipril <- if_else(ukb$ramipril_prevalent == 1 & ukb$amlodipine_prevalent == 0, 0, NA)
ukb$amlodipine_or_ramipril <- if_else(ukb$ramipril_prevalent == 0 & ukb$amlodipine_prevalent == 1, 1, ukb$amlodipine_or_ramipril)
ukb$amlodipine_or_ramipril <- if_else(ukb$ramipril_prevalent == 1 & ukb$amlodipine_prevalent == 1, 2, ukb$amlodipine_or_ramipril)
```

You should get numbers similar to this:

```{r}
table(ukb$amlodipine_or_ramipril)
```


## Hypothesis 1: antihypertensives affect blood pressure

Antihypertensives are prescribed to reduce risk of cardiovascular disease and strokes by lowering blood pressure. Systolic (or "peak") blood pressure is most commonly used clinically to define hypertension.

Test associations between each prevalent antihypertensive and systolic blood pressure separately using linear regression models. Adjust for age and sex.

```{r}
# For example: the association between "previously prescribed amlodipine" and measured SBP
model1 <- glm(systolic_bp ~ age + sex + amlodipine_prevalent, data = ukb)

# extract the model estimates into a data frame
# add variables for the confidence intervals
result1 <- model1 |> summary() |> coef() |> as.data.frame()
result1$CI_lower <- result1$Estimate - (result1$`Std. Error` * 1.96)
result1$CI_upper <- result1$Estimate + (result1$`Std. Error` * 1.96)
result1[-1,]
```

Remember that here, participants with amlodipine are compared to everyone else in the cohort.

Next, test whether SBP is significantly different between patients prescribed amlodipine vs. ramipril. **Hint**: treat the new variable as factor.

```{r}
# For example: the association between "previously prescribed amlodipine" and measured SBP
model2 <- glm(systolic_bp ~ age + sex + as.factor(amlodipine_or_ramipril), data = ukb)

# extract the model estimates into a data frame
# add variables for the confidence intervals
result2 <- model2 |> summary() |> coef() |> as.data.frame()
result2$CI_lower <- result2$Estimate - (result2$`Std. Error` * 1.96)
result2$CI_upper <- result2$Estimate + (result2$`Std. Error` * 1.96)
result2[-1,]
```

Now we are comparing patients prescribed amlodipine (coded as `1`) to those prescribed ramipril (coded as `0`). In the same model we also compare those prescribed both (coded as `2`) to ramipril.


## Hypothesis 2: GP choice of CCB or ACEi depends on patient characteristics

We know from the NICE guidelines that GPs have a choice of medicines to prescribe to a patient following their diagnosis of hypertension.

### Type-2 diabetes

Which medicine do you hypothesise is more common in patients with diagnosed type-2 diabetes? See variables `t2d_prevalent` and `t2d_incident`.

```{r}
# cross-tabulate
table(ukb$t2d_prevalent, ukb$amlodipine_prevalent)
table(ukb$t2d_prevalent, ukb$ramipril_prevalent)
table(ukb$t2d_prevalent, ukb$amlodipine_or_ramipril)
```

Test the hypothesis using logistic regression models. 

```{r}
# run logistic regression model - is ramipril prescription more likely than amlodipine
# if the patient has type-2 diabetes?
model3 <- glm(t2d_prevalent ~ age + sex + as.factor(amlodipine_or_ramipril), data = ukb, family = binomial)

# extract model estimates and calculate CIs
# convert estimates to Odds Ratios
result3 <- model3 |> summary() |> coef() |> as.data.frame()
result3$OR <- exp(result3[,"Estimate"])
result3$CI_lower <- exp(result3$Estimate - (result3$`Std. Error` * 1.96))
result3$CI_upper <- exp(result3$Estimate + (result3$`Std. Error` * 1.96))
result3[-1,]
```

Do the results fit your expectations? Remember that an Odds Ratio (OR) < 1 would indicate that the coded group (e.g., amlodipine `1`) is less likely to have the outcome (in this case type-2 diabetes) compared to the reference group (i.e., ramipril `0`).


### Age

Which medicine do you hypothesise is more common in younger patients? 

Summarize the age at baseline assessment in those with previous prescriptions of amlodipine and ramipril. 

```{r}
summary(ukb$age[ukb$amlodipine_prevalent==1])
summary(ukb$age[ukb$ramipril_prevalent==1])
```

Test the hypothesis that participants prescribed amlodipine are older at baseline assessment using a regression model. **Hint**: the model needs to treat `amlodipine_or_ramipril` as factor

```{r}
# For example: the association between "previously prescribed amlodipine" and measured SBP
model4 <- glm(age ~ sex + as.factor(amlodipine_or_ramipril), data = ukb)

# extract the model estimates into a data frame
# add variables for the confidence intervals
result4 <- model4 |> summary() |> coef() |> as.data.frame()
result4$CI_lower <- result4$Estimate - (result4$`Std. Error` * 1.96)
result4$CI_upper <- result4$Estimate + (result4$`Std. Error` * 1.96)
result4
```

The estimate shows the average difference in age (in units of years) between the groups. It the difference statistically significant?

Use a boxplot to visualize the median age between the two groups.

```{r}
ukb |>
  # drop if missing key data
  filter(!is.na(amlodipine_or_ramipril)) |>
  # Provide "nice" labels
  mutate(group = factor(amlodipine_or_ramipril, 
                        levels = c(0, 1, 2), 
                        labels = c("Ramipril", "Amlodipine", "Both"))
         ) |>
  # Use ggplot to create the boxplot
  ggplot(aes(x = group, y = age, fill = group)) +
  geom_boxplot() +
  labs(
    title = "Age at baseline by antihypertensive medication",
    x = "Antihypertensive medication",
    y = "Age at assessment (years)"
  ) +
  theme(legend.position = "none")
```


# Summary

Congratulations, you've investigated an example of real-world stratified medicine, where patient treatment is tailored to their characteristics. 


```{r, include=FALSE}
## tidy up
system("rm bnf_amlodipine.tsv bnf_ramipril.tsv ukb_gp_scripts.amlodipine.tsv ukb_gp_scripts.ramipril.tsv")
```
